\speculake: Hiding Malware in Speculative Execution

### 2018-08-07 paper/discussion.tex
discussion pass

\paragraph{Indirect Branch Predictor Barrier} - IBPB is used when transitioning to a new
address space, allowing a
program to ensure that earlier code's behavior does not effect its branch prediction.
IBPB requires CPU and operating system support. However, we observe that
processes running under the same user group do not receive IBPB protection,
enabling \speculake when the trigger and payload run under the same group.
Furthermore, IBPB does not prevent the speculative buffer overflow variant of
\speculake described in Section~\ref{subsec:spec-buffer-overflow}.

@@

Finally, \speculake could also use another side channel method that
avoids the cache
to exfiltrate information from the speculative gadgets, such as the branch
predictor itself~\cite{evtyushkin2018branchscope}, memory bandwidth, power
utilization, or contention over other shared resources. While cache 
channels tend to have the highest throughput, they are not the only 
resource that must be monitored to detect or prevent these types of attacks.

%\paragraph{Red-Pills}
%Hardware minutia have also been used to identify and fingerprint execution
%environments---in malware this is known as a red-pill~\cite{red-pill} and is
%typically used by malware to detect when it is under emulation or a debugging
%environment~\cite{lindorfer2011detecting, balzarotti2010efficient,
%paleari2009fistful} attempting to circumvent malware detectors. To address
%red-pills, systems often employ measures to detect them.
\paragraph{Anti-Virus Detectors}
We verified that numerous modern Anti-Virus technologies were unable to detect
and flag \speculake malware. We examined ClamAV, BitDefender and rkhunter and
found that they focus on signature and string detection, examining if potential
malware make library calls, though BitDefender does use their own proprietary
adaptable unpacker to support detection of ``generic
unpacking''~\cite{BitDefender}. Modern anti-virus detectors but do not flag
programs that directly make syscalls or have the capability of making arbitrary
syscalls. We note that this is the state of modern commercial anti-virus
technology but does not mean that \speculake is undetectable.

\paragraph{Bare Metal}
Modern malware often uses hardware minutia to identify and fingerprint execution
environments in order to detect when it is under debugging or
inspection~\cite{lindorfer2011detecting,balzarotti2010efficient,paleari2009fistful}.
To prevent such identification, analyzers often employ ``bare metal''
execution~\cite{kirat2011barebox}, by running the malware on
dedicated hardware that allows introspection and observation of the system
without interfering with its normal operation. This prevents malware from using
so-called ``red pill'' checks to observe that it is under test (and hide its
malicious behavior)~\cite{kirat2014barecloud}. However, to the best of our
knowledge, no publicly available bare-metal environments allow introspection on
the speculative state of the CPU, making it difficult to analyze \speculake
malware. Such environments would be useful for observing the behavior of
\speculake malware in the presence of its trigger, but this presumably would not be
present in such a context.

Symbolic execution has been used to find
environmental red pill checks~\cite{schwartz2010all}. However, such analysis
would be ineffective against \speculake, as symbolic execution does not reason
about speculative paths and how they might influence a program.

@@

\subsubsection{Reverse-engineering triggers}
Once \speculake malware has been detected analysts attempting to
reverse-engineer the effect of the malware can take a similar path as to its
authors. Specifically the task of examining the real-world execution path of the
malware and comparing this with the indirect and direct jumps conducted by
potential trigger programs to find a match. We note that while there may be a
large number of potential trigger programs the real world execution pathway must
match the malware's very closely in order to influence the malware successfully.
However this method will not be useful against the speculative store bypass
version of \speculake, as a secondary trigger program is unnecessary there.

@@

This work demonstrates a general model for hiding execution in 
the speculative world and examines the implications and limitations on 
Intel processors. Given the wide-spread nature of the Spectre vulnerability 
and the ubiquity of side-channels, we believe that this work can 
be directly extended to ARM and other processors making use of
speculative branch prediction.

\subsubsection{Multiple Triggers}
To create further difficulties for an analyst, or to further target the
execution environment, it is possible to have the payload program test a series
of separate jump patterns. Instead of requiring only a single trigger program,
the payload could require multiple trigger programs to be running
simultaneously, or in a particular order.

@@

Alternatively multiple triggers could be used make individual state changes in
the payload program, and write in data from a local, or (as described in
Section~\ref{subsec:openssl}) remote processes. For instance, triggers could be
used to shift in data or an entire program for emulation.
%For example, a payload program
%can check a series of three distinct jump patterns corresponding to three
%different triggers. If the first trigger is detected the payload shifts in a 1
%to the state; if the second is detected it shifts in a 0. If the third trigger
%is detected it discontinues listening to the three triggers and executes the
%data shifted into the state.
%This could be a direct form of \speculake using emulation and
%decryption or more traditional malicious behavior. 
Using this method a payload
program can be distributed without the encrypted data section all together,
instead having the data written into the process at run-time without ever
connecting out to any other resource.

@@

\subsubsection{Hypervisors}
For virtual environments that provide guests significant periods of access to
hardware resources, \speculake is viable in certain configurations. 
This result is specific to Spectre 2 variant, as it relies on 
the indirect branch predictor effecting the branches in either the host 
or guest environment. We have found that the hypercall context switch from 
guest to host is light enough that a trigger program running in a guest
can effect the branch predictor in a host on the same CPU core. This is not
true in the opposite direction (where the trigger runs on the host) as the 
hypercall transition seems to be more substantial. While this implies 
that guest-to-guest indirect branch predictor interference, and the current
\speculake implementation, will not work across disparate virtual environments,
strict resource isolation gives a stronger guarantee of non-interference.

\subsubsection{Alternative Applications}
We have proposed the \speculake model as a new malware threat, however, this
model of hidden computation applies more broadly. As of now there are no ways of
debugging speculative computation, meaning that the computation done there is
effectively isolated and difficult to extract by third-parties. While not a true
trusted execution environment, it could be used like one, offering potential
uses in Digitial Rights Management (DRM), and whitebox cryptography.

>>>>

\paragraph{Indirect Branch Predictor Barrier} - IBPB is used when transitioning to a new
address space, allowing a
program to ensure that earlier code's behavior does not effect its branch prediction.
IBPB requires CPU and operating system support. However, we observe on Linux that
processes running under the same user group do not receive IBPB protection,
enabling \speculake when the trigger and payload run under the same group.
Furthermore, IBPB does not prevent the speculative buffer overflow variant of
\speculake described in Section~\ref{subsec:spec-buffer-overflow}.

@@

\speculake could also use another side channel method that
avoids the cache
to exfiltrate information from the speculative gadgets, such as the branch
predictor itself~\cite{evtyushkin2018branchscope}, memory bandwidth, power
utilization, or contention over other shared resources~\cite{kiriansky2018speculative}.
While cache 
channels tend to have the highest throughput, they are not the only 
resource that must be monitored to detect or prevent these types of attacks.

%\paragraph{Red-Pills}
%Hardware minutia have also been used to identify and fingerprint execution
%environments---in malware this is known as a red-pill~\cite{red-pill} and is
%typically used by malware to detect when it is under emulation or a debugging
%environment~\cite{lindorfer2011detecting, balzarotti2010efficient,
%paleari2009fistful} attempting to circumvent malware detectors. To address
%red-pills, systems often employ measures to detect them.
\paragraph{Anti-Virus Detectors}
We verified that modern Anti-Virus technologies were unable to detect
and flag \speculake malware. We used ClamAV, BitDefender and rkhunter, which
mainly rely on signature- and string-based detection. BitDefender does feature
support for unpacking or extracting malware, though appears to simply try
unpacking using several known packers and encoding formats~\cite{BitDefender}.
Thus, it is not surprising that these tools cannot detect
\speculake.

\paragraph{Bare Metal}
Modern malware often uses hardware minutia to identify and fingerprint execution
environments in order to detect when it is under debugging or
inspection~\cite{lindorfer2011detecting,balzarotti2010efficient,paleari2009fistful}.
To prevent such identification, analyzers often employ ``bare metal''
execution~\cite{kirat2011barebox}, running the malware on
dedicated hardware that allows introspection and observation of the system
without interfering with its normal operation. This prevents malware from using
so-called ``red pill'' checks to observe that it is under test (and hide its
malicious behavior)~\cite{kirat2014barecloud}. However, to the best of our
knowledge, no publicly available bare-metal environments allow introspection on
the speculative state of the CPU, making it difficult to analyze \speculake
malware. However, such environments could be useful for observing the behavior of
\speculake malware in the presence of its trigger if available, as modifications
in the real world could be easily tracked.

%but this presumably would not be
%present in such a context.

Symbolic execution has also been used to find
environmental red pill checks~\cite{schwartz2010all}. However, such analysis
would be ineffective against \speculake, as symbolic execution does not reason
about speculative paths and how they might influence a program.

@@

\subsubsection{Reverse-engineering triggers}
For program-based triggers, an analyst could attempt to find the trigger program
by examining the execution path of the payload program, and locating a common
indirect jump pattern between payload and potential triggers. Since both programs must
share a common indirect jump pattern to interact via the indirect branch
predictor, there must be some overlap which is unlikely to occur randomly
between two programs.

We note that while the analyst may learn the execution path (and thus true
indirect jump pattern) of the payload program, they may not be able to capture
every potential execution path in all potential triggers. For example, in the
OpenSSL trigger, the analyst may not have captured all potential indirect jump
patterns, as doing so would require exhaustively connecting to OpenSSL with
different cipher suites, extensions, and failed handshakes. However, the analyst
can still make a list of indirect jump locations in a suspected trigger program,
comparing these to the jumps taken by the payload. If there is significant
overlap, the analyst could spend time to discover what inputs to the trigger
program produce similar indirect jump patterns, thus discovering the trigger.

\speculake malware could attempt to thwart this analysis by using decoy indirect
jumps that do not correspond with the trigger, but potentially correspond with
other (non-trigger) binaries. In addition, this analysis method is ineffective
at inspecting the speculative buffer overflow variant described in
Section~\ref{subsec:spec-buffer-overflow}, as it does not use a separate trigger
program.

@@

\speculake demonstrates a general model for hiding execution in 
the speculative world and examines the implications and limitations on
modern processors. Given the wide-spread nature of the Spectre vulnerability 
and the ubiquity of side-channels, we believe that this work can 
be directly extended to other architectures, such as ARM, and other processors
making use of
speculative branch prediction.

\subsubsection{Multiple Triggers}
To create further difficulties for an analyst, or to further target the
execution environment, it is possible to have the payload program to combine
multiple triggers. Instead of requiring only a single trigger program,
the payload could require multiple trigger programs to be running
simultaneously, or in a particular order. Alternatively, the payload could combine
trigger programs with input triggers, forcing an analyst to understand multiple
variants simulatenously.

@@

Alternatively multiple triggers could be used make individual state changes in
the payload program, and write in data from a local, or (as described in
Section~\ref{subsec:openssl}) remote processes. For instance, triggers could be
used to shift in data or an entire program for emulation.
%For example, a payload program
%can check a series of three distinct jump patterns corresponding to three
%different triggers. If the first trigger is detected the payload shifts in a 1
%to the state; if the second is detected it shifts in a 0. If the third trigger
%is detected it discontinues listening to the three triggers and executes the
%data shifted into the state.
%This could be a direct form of \speculake using emulation and
%decryption or more traditional malicious behavior. 
Using this method a payload
program can be distributed without the encrypted data section at all,
instead having the data written into the process indirectly at run-time without ever
connecting out to any other resource.

@@

\subsubsection{Virtual Machines}
Virtual environments could also be host to \speculake malware and triggers. For
instance, malware on one EC2 instance could potentially be triggered by a
trigger program on another seemingly unrelated instance. We have found that the
hypercall context switch from guest to host on VirtualBox is lightweight enough
that a trigger program running in a guest can activate a payload program running
in the host on the same CPU core. However, we have so far been unable to go in
the opposite direction, and similarly have yet to achieve guest-to-guest
interference. More work is needed to determine if such barriers are possible to
overcome, and if stronger isolation is needed in the virtual machine context.

%For virtual environments that provide guests significant periods of access to
%hardware resources, \speculake is viable in certain configurations. 
%This result is specific to Spectre 2 variant, as it relies on 
%the indirect branch predictor effecting the branches in either the host 
%or guest environment. We have found that the hypercall context switch from 
%guest to host is light enough that a trigger program running in a guest
%can effect the branch predictor in a host on the same CPU core. This is not
%true in the opposite direction (where the trigger runs on the host) as the 
%hypercall transition seems to be more substantial. While this implies 
%that guest-to-guest indirect branch predictor interference, and the current
%\speculake implementation, will not work across disparate virtual environments,
%strict resource isolation gives a stronger guarantee of non-interference.

\subsubsection{Alternative Applications}
We have proposed the \speculake model as a new malware threat, however, this
model of hidden computation could be applied more broadly. As of now there are no ways of
debugging speculative computation, meaning that the computation done there is
effectively isolated and difficult to extract by third-parties, even with
physical or root access to the machine it operates on. While not a true
trusted execution environment, it is possible that such an environment could be
used as a limited trusted environment. This may have potential use in digital
rights management (DRM), and whitebox cryptography applications.
